---
title: "Multiome Processing and Clustering"
author: "Jillian Poska"
date: "2026-06-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries
```{r}
library(hdf5r)
library(Signac)
library(Seurat)
library(SeuratObject)
library(tidyverse)
library(dplyr)
library(umap)
library(glmGamPoi)

library(SingleCellExperiment)
library(SeuratWrappers)
library(SoupX)


setwd("~/Multiome2")
set.seed(228)
```

## Process and normalize each sample individually
SoupX works best with some kind of clustering information already provided. I processed each sample individually to obtain the clustering information to run SoupX. This gave me a rough idea, but I reclustered my data later on, so don't worry if it's not perfect.
```{r}
# Load data
BPA.counts <- Read10X_h5("~/Multiome2/BPA Adult/filtered_feature_bc_matrix.h5")
BPA.fragments <- "~/Multiome2/BPA Adult/atac_fragments.tsv.gz"
oil.counts <- Read10X_h5("~/Multiome2/Veh Adult/filtered_feature_bc_matrix.h5")
oil.fragments <- "~/Multiome2/Veh Adult/atac_fragments.tsv.gz"

# Create Seurat object using RNA counts
BPA <- CreateSeuratObject(
  counts = BPA.counts$`Gene Expression`,
  assay = "RNA"
)
oil <- CreateSeuratObject(
  counts = oil.counts$`Gene Expression`,
  assay = "RNA"
)

library(EnsDb.Mmusculus.v79)
# get gene annotations from mm10
annotations <- GetGRangesFromEnsDb(ensdb = EnsDb.Mmusculus.v79)
seqlevels(annotations) <- paste0('chr',seqlevels(annotations))

# create ATAC assay and add to the Seurat object
BPA[["ATAC"]] <- CreateChromatinAssay(
  counts = BPA.counts$Peaks,
  sep = c(":","-"),
  fragments = BPA.fragments,
  annotation = annotations
)

oil[["ATAC"]] <- CreateChromatinAssay(
  counts = oil.counts$Peaks,
  sep = c(":","-"),
  fragments = oil.fragments,
  annotation = annotations
)

# Normalize and cluster each sample
# BPA
DefaultAssay(BPA) <- "RNA"

BPA <- BPA %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  RunUMAP(dims = 1:20, reduction.name = 'umap.rna', reduction.key = 'rnaUMAP_')

DefaultAssay(BPA) <- "ATAC"
BPA <- BPA %>%
  RunTFIDF() %>%
  FindTopFeatures(min.cutoff = 'q0') %>%
  RunSVD() %>%
  RunUMAP(reduction = "lsi", dims = 2:50, reduction.name = "umap.atac", reduction.key = "atacUMAP_")

BPA <- FindMultiModalNeighbors(BPA, reduction.list = list("pca", "lsi"), dims.list = list(1:20, 2:50))
BPA <- RunUMAP(BPA, nn.name = "weighted.nn", reduction.name = "wnn.umap", reduction.key = "wnnUMAP_")
BPA <- FindClusters(BPA, graph.name = "wsnn", algorithm = 3, verbose = FALSE, resolution = 0.2)

# oil
DefaultAssay(oil) <- "RNA"

oil <- oil %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  RunUMAP(dims = 1:20, reduction.name = 'umap.rna', reduction.key = 'rnaUMAP_')

DefaultAssay(oil) <- "ATAC"
oil <- oil %>%
  RunTFIDF() %>%
  FindTopFeatures(min.cutoff = 'q0') %>%
  RunSVD() %>%
  RunUMAP(reduction = "lsi", dims = 2:50, reduction.name = "umap.atac", reduction.key = "atacUMAP_")

oil <- FindMultiModalNeighbors(oil, reduction.list = list("pca", "lsi"), dims.list = list(1:20, 2:50))
oil <- RunUMAP(oil, nn.name = "weighted.nn", reduction.name = "wnn.umap", reduction.key = "wnnUMAP_")
oil <- FindClusters(oil, graph.name = "wsnn", algorithm = 3, verbose = FALSE, resolution = 0.2)

# Save metadata
BPA_metadata <- BPA@meta.data
BPA_umap <- BPA@reductions$wnn.umap
BPA_embeddings <- BPA_umap@cell.embeddings
remove(BPA_umap)

oil_metadata <- oil@meta.data
oil_umap <- oil@reductions$wnn.umap
oil_embeddings <- oil_umap@cell.embeddings
remove(oil_umap)

remove(BPA, oil)
```

## Run SoupX analysis to remove ambient RNA contamination
```{r}
# Upload raw feature matrices for each sample and make a soup channel
BPA.toc <- BPA.counts$`Gene Expression`
BPA.tod <- Read10X_h5("BPA Adult/raw_feature_bc_matrix.h5")
BPA.tod <- BPA.tod$`Gene Expression`
BPA.sc <- SoupChannel(toc = BPA.toc, tod = BPA.tod)

oil.toc <- oil.counts$`Gene Expression`
oil.tod <- Read10X_h5("Veh Adult/raw_feature_bc_matrix.h5")
oil.tod <- oil.tod$`Gene Expression`
oil.sc <- SoupChannel(toc = oil.toc, tod = oil.tod)

remove(BPA.toc, BPA.tod, oil.toc, oil.tod)

# Enter clustering information and embeddings into the soup channel
BPA.sc <- setClusters(BPA.sc, setNames(BPA_metadata$seurat_clusters, rownames(BPA_metadata)))
BPA.sc <- setDR(BPA.sc, BPA_embeddings[colnames(BPA.sc$toc), c("wnnUMAP_1", "wnnUMAP_2")])

oil.sc <- setClusters(oil.sc, setNames(oil_metadata$seurat_clusters, rownames(oil_metadata)))
oil.sc <- setDR(oil.sc, oil_embeddings[colnames(oil.sc$toc), c("wnnUMAP_1", "wnnUMAP_2")])
remove(BPA_metadata, BPA_embeddings, oil_metadata, oil_embeddings)

# Set contamination fraction - based on auto estimates and using top soup genes to estimate
BPA.sc <- setContaminationFraction(BPA.sc, 0.2)
oil.sc <- setContaminationFraction(oil.sc, 0.05)

# Correct expression levels
BPA.out <- adjustCounts(BPA.sc)
oil.out <- adjustCounts(oil.sc)

# Make a Seurat object with the corrected RNA counts
BPA <- CreateSeuratObject(BPA.out,
                          assay = "RNA")

oil <- CreateSeuratObject(oil.out,
                          assay = "RNA")
remove(oil.sc, oil.out, BPA.sc, BPA.out)
```

## Add ATAC data to Seurat object
```{r}
bpa.atac <- BPA.counts$Peaks
oil.atac <- oil.counts$Peaks

# Limit analysis to only standard chromosomes
bpa.atac.grange <- StringToGRanges(rownames(bpa.atac), sep = c(":", "-"))
bpa.grange.use <- seqnames(bpa.atac.grange) %in% standardChromosomes(bpa.atac.grange)
bpa.atac <- bpa.atac[as.vector(bpa.grange.use), ]

oil.atac.grange <- StringToGRanges(rownames(oil.atac), sep = c(":", "-"))
oil.grange.use <- seqnames(oil.atac.grange) %in% standardChromosomes(oil.atac.grange)
oil.atac <- oil.atac[as.vector(oil.grange.use), ]

# create ATAC assay and add to the Seurat object
BPA[["ATAC"]] <- CreateChromatinAssay(
  counts = bpa.atac,
  sep = c(":","-"),
  fragments = BPA.fragments,
  annotation = annotations,
  genome = "mm10"
)

oil[["ATAC"]] <- CreateChromatinAssay(
  counts = oil.atac,
  sep = c(":","-"),
  fragments = oil.fragments,
  annotation = annotations,
  genome = "mm10"
)

remove(BPA.counts, oil.counts, BPA.fragments, oil.fragments, annotations, bpa.atac.grange, bpa.atac, bpa.grange.use, oil.atac, oil.atac.grange, oil.grange.use)
```

## QC data
```{r}
# set default assay to ATAC and calculate QC parameters
DefaultAssay(BPA) <- "ATAC"

BPA <- NucleosomeSignal(BPA)
BPA <- TSSEnrichment(BPA)

DefaultAssay(oil) <- "ATAC"

oil <- NucleosomeSignal(oil)
oil <- TSSEnrichment(oil)

# check QC metrics for ATAC
DensityScatter(BPA, x = "nCount_ATAC", y = "TSS.enrichment", log_x = TRUE, quantiles = TRUE)

VlnPlot(
  object = BPA,
  features = c( "nCount_ATAC", "TSS.enrichment", "nucleosome_signal"),
  ncol = 4,
  pt.size = 0
)

DensityScatter(oil, x = "nCount_ATAC", y = "TSS.enrichment", log_x = TRUE, quantiles = TRUE)

VlnPlot(
  object = oil,
  features = c( "nCount_ATAC", "TSS.enrichment", "nucleosome_signal"),
  ncol = 4,
  pt.size = 0
)

# calculate QC metrics for RNA
DefaultAssay(BPA) <- "RNA"
# Soup genes as determined from the SoupProfile
soup.genes <- c("Csn2", "Csn1s2a", "Glycam1", "Tpt1", "Csn3", "Spp1", "Camk1d", "Wap", "Fabp4", "AY036118")

BPA[["percent.soup"]] <- PercentageFeatureSet(BPA, features = soup.genes)
BPA[["percent.mt"]] <- PercentageFeatureSet(BPA, pattern = "^mt-")

VlnPlot(BPA, features = c("percent.mt", "nCount_RNA", "percent.soup"), ncol = 4, pt.size = 0)

DefaultAssay(oil) <- "RNA"

oil[["percent.mt"]] <- PercentageFeatureSet(oil, pattern = "^mt-")

VlnPlot(oil, features = c("percent.mt", "nCount_RNA"), ncol = 3, pt.size = 0)

# subset data to exclude cells failing QC metrics
BPA <- subset(
  x = BPA,
  subset = nCount_ATAC < 50000 &
    nCount_RNA < 20000 &
    nCount_ATAC > 2000 &
    nCount_RNA > 1200 &
    nucleosome_signal < 2 &
    TSS.enrichment > 4 &
    percent.mt < 20 &
    percent.soup < 5
)

oil <- subset(
  x = oil,
  subset = nCount_ATAC < 50000 &
    nCount_RNA < 20000 &
    nCount_ATAC > 1000 &
    nCount_RNA > 1000 &
    nucleosome_signal < 2 &
    TSS.enrichment > 3 &
    percent.mt < 20
)

remove(BPA.dc.sobj, oil.dc.sobj)
```

## Merge data
Credit to the QuadBio multiome vignette for the code for creating a union peak list
```{r}
BPA@meta.data$treatment <- "BPA"
oil@meta.data$treatment <- "Vehicle"

# merge datasets
adult <- merge(
  x = BPA,
  y = oil,
  add.cell.ids = c("BPA", "oil")
)

# create a union peak list 
peaks <- GenomicRanges::reduce(unlist(as(c(BPA@assays$ATAC@ranges,
                                           oil@assays$ATAC@ranges),
                                         "GRangesList")))

peakwidths <- width(peaks)
peaks <- peaks[peakwidths < 10000 & peakwidths > 20]

counts_atac_merged <- FeatureMatrix(adult@assays$ATAC@fragments,
                                    features = peaks,
                                    cells = colnames(adult))

# add peak list to merged Seurat
adult[['ATAC']] <- CreateChromatinAssay(counts_atac_merged,
                                               fragments = adult@assays$ATAC@fragments,
                                               annotation = adult@assays$ATAC@annotation,
                                               sep = c(":","-"))

remove(BPA, oil)

```

## Processing data
```{r}
DefaultAssay(adult) <- "RNA"

adult <- adult %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA() %>%
  RunUMAP(dims = 1:50, reduction.name = 'umap.rna', reduction.key = 'rnaUMAP_')

adult <- IntegrateLayers(object = adult, 
    method = HarmonyIntegration,
    orig.reduction = "pca", 
    new.reduction = "integrated.harmony",
    verbose = TRUE)

DefaultAssay(adult) <- "ATAC"
adult <- RunTFIDF(adult)
adult <- FindTopFeatures(adult, min.cutoff = 'q0')
adult <- RunSVD(adult)
adult <- RunUMAP(adult, reduction = 'lsi', dims = 2:50, reduction.name = "umap.atac", reduction.key = "atacUMAP_")

integration.anchors <- FindIntegrationAnchors(
  object.list = SplitObject(adult, "treatment"),
  anchor.features = rownames(adult),
  reduction = "rlsi",
  dims = 2:50
)
 
adult_integrated <- IntegrateEmbeddings(
  anchorset = integration.anchors,
  reductions = adult[["lsi"]],
  new.reduction.name = "integrated.lsi",
  dims.to.integrate = 1:50
)

adult[['integrated_lsi_atac']] <- CreateDimReducObject(
  Embeddings(adult_integrated, "integrated.lsi")[colnames(adult),], key = "INTEGRATEDLSI_", assay = "ATAC"
)
remove(integration.anchors, adult_integrated)
```

## Clustering
```{r}
adult <- FindMultiModalNeighbors(adult, reduction.list = list("integrated.harmony", "integrated_lsi_atac"), dims.list = list(1:50, 2:50))
adult <- RunUMAP(adult, nn.name = "weighted.nn", reduction.name = "wnn.umap", reduction.key = "wnnUMAP_")
adult <- FindClusters(adult, graph.name = "wsnn", algorithm = 3, verbose = FALSE, resolution = 0.2)

DefaultAssay(adult) <- "RNA"
adult[["RNA"]] <- JoinLayers(adult[["RNA"]])
rna_markers <- FindAllMarkers(object = adult,
                          min.pct = 0.3,
                          only.pos = T,
                          logfc.threshold = 0.3)
rna_markers <- rna_markers[rna_markers$p_val_adj < 0.05, ]

# Supplemental Figure 
DotPlot(adult, features = c("Epcam", "Elf5", "Wnt5b", "Esr1", "Cited1", "Krt14", "Acta2", "Lgr5", "Pdgfra", "Col1a1", "Pi16", "Rgs5", "Notch3", "Pecam1", "Cdh5", "Prox1", "Mmrn1", "Ptprc", "Csf1r", "F13a1", "Cd3d", "Cd3e", "Cd3g", "Cd79a", "Cd79b"))
```

## Check if there is any difference between separate cell type clusters
```{r}
DefaultAssay(adult) <- "RNA"
basal.de.markers <- FindMarkers(adult, ident.1 = "Basal 1", ident.2 = "Basal 2")
basal1.markers <- basal.de.markers[basal.de.markers$avg_log2FC > 0.5, ]
basal2.markers <- basal.de.markers[basal.de.markers$avg_log2FC < -0.5, ]

tcell.de.markers <- FindMarkers(adult, ident.1 = "T cell 1", ident.2 = "T cell 2")
tcell1.markers <- tcell.de.markers[tcell.de.markers$avg_log2FC > 0.5, ]
tcell2.markers <- tcell.de.markers[tcell.de.markers$avg_log2FC < - 0.5, ]

le.de.markers <- FindMarkers(adult, ident.1 = "Lymphatic Endothelial 1", ident.2 = "Lymphatic Endothelial 2")
le1.markers <- le.de.markers[le.de.markers$avg_log2FC > 0.5, ]
le2.markers <- le.de.markers[le.de.markers$avg_log2FC < -0.5, ]

ve.de.markers <- FindMarkers(adult, ident.1 = "Vascular Endothelial 1", ident.2 = "Vascular Endothelial 2")
ve1.markers <- ve.de.markers[ve.de.markers$avg_log2FC > 0.5, ]
ve2.markers <- ve.de.markers[ve.de.markers$avg_log2FC < -0.5, ]
```


## Rename clusters based on RNA markers
```{r}
Idents(adult) <- "seurat_clusters"
adult <- RenameIdents(object = adult, '0' = "Mature Luminal")
adult <- RenameIdents(object = adult, '1' = "Vascular Endothelial 1")
adult <- RenameIdents(object = adult, '2' = "Basal 1")
adult <- RenameIdents(object = adult, '3' = "Fibroblast 1")
adult <- RenameIdents(object = adult, '4' = "Luminal Progenitor")
adult <- RenameIdents(object = adult, '5' = "Macrophage 1")
adult <- RenameIdents(object = adult, '6' = "Fibroblast 2")
adult <- RenameIdents(object = adult, '7' = "Basal 2")
adult <- RenameIdents(object = adult, '8' = "Vascular Endothelial 2")
adult <- RenameIdents(object = adult, '9' = "T cell 1")
adult <- RenameIdents(object = adult, '10' = "Lymphatic Endothelial 1")
adult <- RenameIdents(object = adult, '11' = "Pericyte")
adult <- RenameIdents(object = adult, '12' = "Macrophage 2")
adult <- RenameIdents(object = adult, '13' = "T cell 2")
adult <- RenameIdents(object = adult, '14' = "Lymphatic Endothelial 2")
adult <- RenameIdents(object = adult, '15' = "B cell")
adult <- RenameIdents(object = adult, '16' = "Fibroblast 3")


adult$celltype <- adult@active.ident
```
